---
title: "Sequence a genome II"
author: "Kamil S Jaroň"
date: "Spring 2016"
output: ioslides_presentation
subtitle: mapping
---

## This semester

- 24 . - 26.02.  : RNA sequencing (done)
- 12.04. : Data exploration
- 19.04. : **Mapping reads and quantifying**
- 26.04. : Downstream analyses
- 03.05. : Downstream analyses
- 10.05. : Analysis of annotated genes
- 17.05. : Analysis of annotated genes
- 24.05. : Presentations

## Today

- 12.04. : Mapping reads and quantifying

    - indexing of cds (.fasta -> .idx)
    - mapping of reads (.idx and .fastq -> .sam and .tsv)
    - RNAseq theory
    - visualisation of mapping (IGV)

## Vital-it reminder

Long story short

`/home/<username>` - scripts, configuration files

`/scratch/cluster/monthly/<username>` - data files, analyses

`prd` - is a computer you are logging to, the interface between you and cluster
`bsub` - is a command you use to run a task on cluster, not on the `prd` computer

## Organisation of our files

Compute everything in **your** folder. 

When the job is done, move the result to common folders in `/scratch/cluster/monthly/mls_2015/SAGE/...`

>- `RNAseq/` - raw reads (fastq.gz)
>- `genome/` - genome and annotatation (.fasta, .gtf, .gff)
>- `RNAseq_trimmed/` - prepared for trimmed reads (.fq.gz)
>- `scripts/` - all scripts and script templetes
>- `mapping/` - prepared for mapping files (.bam, .bam.bai)
>- `read_counts/` - prepared for read counts

## Your turn...

MyUnil > courses > Sequence a Genome II > RNAseq2_mapping.html

## So to keep everything in order...

Check a quality control of reads of yours...

If your trimmed reads are fine, move them to appropriate folder

```bash
mv <my_trimmed_reads.fastq.gz> \
/scratch/cluster/monthly/mls_2015/SAGE/RNAseq_trimmed/<condition>/
```

## packagess of Vital-it

```bash
module add <package_path>/<package_name>
```

## Indexing of coding regions

```bash
bsub -q priority -J <job_name> \
    kallisto index -i <indexed_cds.idx> <S5_cds.fasta>
```

- Parameter `-q` of the `bsub` set a queue for your job

- `<indexed_cds.idx>` will be the output
- `<S5_cds.fasta>` is the file with coding regions we created last week

## Kallisto output

```bash
kallisto quant -i <indexed_cds.idx> -o <output_folder> -b 100 --single \
              -l <mean> -s <sd> --pseudobam <input_files> > <output.sam>
```

`./<output_folder>/`

 - `abundances.h5` is a HDF5 binary file containing run info, abundance estimates, bootstrap estimates, and transcript length information length
 - `abundances.tsv` is a plain text file of the abundance estimates.
 - `run_info.json` is a file containing information about the run

## Kallisto output

```bash
kallisto quant -i <indexed_cds.idx> -o <output_folder> -b 100 \
  --single -l <mean> -s <sd> --pseudobam <input_files> \
  > <output.sam>
```

`./<output.sam>` is a file with pseudoalignment

## Break

...on through to the other side

##{}

```{r fig.width=8, fig.height=6,echo=FALSE}
library(png)
library(grid)
img <- readPNG("./bow_kal_sam_header.png")
 grid.raster(img)
```

##{}

```{r fig.width=8, fig.height=6,echo=FALSE}
library(png)
library(grid)
img <- readPNG("./bow_kal_sam_alignment.png")
 grid.raster(img)
```

##A script

```bash
#!/bin/bash

#BSUB -L /bin/bash
#BSUB -J <job_name>
#BSUB -q normal

module add UHTS/Analysis/samtools/latest

python3 kallisto_sam_convertor.py <pseudoalignment.sam> \
  /scratch/.../SAGE/genome/Pseudomonas_protegens_S5_genome.gtf \
  | samtools view -bS - | samtools sort - -o <output.bam>
```

##How to edit a script

```bash
cd /scratch/cluster/monthly/mls_2015/<username>/<kallisto_folder>
cp ../../SAGE/scripts/* .
```

```bash
vim submit_conversion.sh
```

##Vim - opened file

```{r fig.width=8, fig.height=6,echo=FALSE}
library(png)
library(grid)
img <- readPNG("./vim1.png")
 grid.raster(img)
```

##Vim - insert mode

```{r fig.width=8, fig.height=6,echo=FALSE}
library(png)
library(grid)
img <- readPNG("./vim2.png")
 grid.raster(img)
```

##Vim - commad mode, quit and write

```{r fig.width=8, fig.height=6,echo=FALSE}
library(png)
library(grid)
img <- readPNG("./vim3.png")
 grid.raster(img)
```

## How to submit a sript to vital-it

```bash
bsub < submit_conversion.sh
```

## IGV

- Find out how to load a data to IGV.
- Does the mapping have a sense?
- Can you check a gene you have annotated last semester?
- Do you see any "expressed" non coding region?
- Is it technically possible to see a read mapped to non coding region, given the fact, we used kallisto?
- Why?

## Scheduled for the next week

  >- Quality accessment (yes, again)
  >- Clustering of sample-to-sample distances
  >- Differential expression
  
## See you next week

Thanks for your participation!